---
title: "Kenya’s Biodiversity in GBIF: The Good, the Gaps, the Path Forward"
author: "Jean K"
date: "2025-12-09"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false

# Load required packages
library(here)
library(dplyr)
library(tidyr)
library(sf)
library(plotly)
library(leaflet)
library(DT)
library(knitr)
library(kableExtra)

# Set paths relative to this article's location (in blog/Gbif Kenya/docs/)
results_dir <- here::here("blog", "Gbif Kenya", "results")
figures_dir <- here::here("blog", "Gbif Kenya", "figures")

# Load pre-processed data (assumed to be prepared in earlier steps)
env_comparison <- readRDS(file.path(results_dir, "spatial_bias", "environmental_comparison.rds"))
taxonomic_summary <- readRDS(here(results_dir, "taxonomic_bias", "taxonomic_bias_summary.rds"))
data_quality <- readRDS(here(results_dir, "data_quality", "data_quality_assessment.rds"))
class_summary <- readRDS(here(results_dir, "taxonomic_bias", "class_summary.rds"))
```

I set out to understand Kenya's biodiversity data. What I discovered was a mirror reflecting how we do science itself.
Kenya hosts extraordinary biodiversity — from montane forests to coral reefs, from the Maasai Mara’s megafauna to thousands of endemic species found nowhere else. But when I downloaded 4.9 million occurrence records from [GBIF](https://www.gbif.org/) to assess what we actually *know*, the picture that emerged was humbling.

::: {.callout-note}
## Scope of this analysis

All the numbers, maps and statistics in this article are based on **open, GBIF-mediated occurrence records for Kenya**. They do **not** include internal monitoring and research data held by institutions such as Kenya Wildlife Service (KWS), the Wildlife Research and Training Institute (WRTI), National Museums of Kenya, universities, NGOs or private conservancies, many of which are **not yet publicly accessible or not fully mobilised to GBIF**.

You can think of this as a view of **Kenya's biodiversity as seen through the global GBIF window**, not a complete inventory of every observation ever collected in the country.
:::

## Discovery 1: We've barely scratched the surface

Only **13.3%** of Kenya's land area contains any recorded biodiversity observations.

Let that sink in. Despite decades of research, expeditions, and citizen science, **86.7%** of the country — nearly 500,000 km² — has zero occurrence records in the world’s largest biodiversity database.

The arid and semi-arid lands covering over 80% of Kenya? Drastically undersampled. Northern Kenya’s unique ecosystems? Virtually invisible. Coastal zones and high-altitude peaks? Significant gaps.

Our biodiversity maps may actually be maps of where researchers travel, not where species live.

### Interactive Map: Where We've Actually Sampled

```{r sampling-map}
#| fig-cap: "Interactive map of sampling effort across Kenya. Hexagonal cells (~10 km resolution) show the number of occurrence records. Click cells for details."

# Example leaflet map using pre-computed hexagon summaries
hex_grid <- readRDS(here(results_dir, "spatial_bias", "spatial_grid_effort.rds"))

pal <- colorNumeric(
  palette = "viridis",
  domain = log10(hex_grid$n_records + 1)
)

leaflet(hex_grid) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(
    fillColor = ~pal(log10(n_records + 1)),
    weight = 0.2,
    opacity = 0.7,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.8,
      bringToFront = TRUE
    ),
    label = ~paste0(
      "Records: ", format(n_records, big.mark = ","),
      "<br>Unique species: ", format(n_species, big.mark = ",")
    ) %>% lapply(htmltools::HTML)
  ) %>%
  addLegend(
    pal = pal,
    values = ~log10(n_records + 1),
    title = "Log10(Records)",
    position = "bottomright",
    opacity = 0.7
  ) %>%
  addControl(
    html = "<div style='background: white; padding: 10px; border-radius: 5px;'>
      <strong>Sampling Effort Map</strong><br>
      <small>• Zoom to explore regions<br>
      • Click cells for statistics<br>
      • Note clustering near cities</small>
    </div>",
    position = "topright"
  )
```

## Discovery 2: "Biodiversity data" is really bird data

When I calculated taxonomic representation, one number stopped me cold: **87.5%**.

That's the proportion of all Kenya's GBIF records belonging to a single class — **Aves (birds)**. The Gini coefficient measuring taxonomic inequality came out at **0.966**, indicating extreme concentration.

Birds are magnificent and deserve study. But Kenya’s insects, fungi, soil microbes, marine invertebrates, and countless other groups collectively comprise far more of actual biodiversity. They’re nearly absent from the open data.

In other words, we’re not measuring Kenya’s biodiversity. We’re measuring what humans find charismatic and convenient to observe.

It’s also worth stressing that this is **a GBIF-specific picture**. National institutions and research projects in Kenya hold additional records for plants, mammals, invertebrates and other groups that aren’t yet fully mobilised to GBIF or are only accessible under specific data-sharing agreements. Those internal datasets are logically separate from the GBIF layer I analyse here.

### Interactive Treemap: What We've Actually Recorded

```{r taxonomic-treemap}
#| fig-cap: "Taxonomic representation in Kenya’s GBIF data. Box size = number of records. Aves (birds) dominate with 87.5% of all records."

# Example treemap using taxonomic_summary data
library(treemapify)
library(ggplot2)

ggplot(
  class_summary,
  aes(
    area = n_records,
    fill = kingdom,
    label = paste(class, "\n", scales::comma(n_records), "records")
  )
) +
  geom_treemap(color = "white") +
  geom_treemap_text(
    place = "centre",
    grow = TRUE,
    reflow = TRUE,
    colour = "white",
    min.size = 2
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    fill = "Kingdom",
    title = "Taxonomic Representation in Kenya's GBIF Data"
  )
```

```{r taxonomic-summary-table}
kable(
  class_summary %>%
    arrange(desc(n_records)) %>%
    mutate(
      Records = scales::comma(n_records),
      Species = scales::comma(n_species),
      `% of Total` = sprintf("%.1f%%", 100 * n_records / sum(class_summary$n_records))
    ) %>%
    select(Kingdom = kingdom, Phylum = phylum, Class = class,
           Records, Species, `% of Total`),
  format = "html",
  caption = "Summary of key taxonomic groups in Kenya's GBIF data.",
  table.attr = "class=\"table table-striped table-hover\""
) %>%
  kable_styling(full_width = FALSE)
```

## Discovery 3: More than half the data has quality issues

**54.7%** of records carried at least one quality flag.

- **42.5%** were duplicates — the same observation submitted through multiple data pathways.
- **12.9%** had coordinate issues, including records mysteriously placed at country centroids or capital cities (a telltale sign of imprecise georeferencing).
- Nearly **7%** lacked species-level identification.

This isn’t a criticism of data collectors. It reflects the realities of aggregating data across institutions, decades, and methodologies. But it means we **cannot treat these records as uniformly reliable**.

Again, these flags apply **only to the GBIF-mediated slice** of Kenya’s biodiversity information. Internal databases (for example, those used for protected area monitoring or research station work) may apply different quality controls and may never expose problematic records to the public GBIF layer.

### Data Quality Flags Overview

```{r data-quality-summary}
#| fig-cap: "Distribution of data quality flags across Kenya’s GBIF occurrence records."
ggplot(data_quality$tracking, aes(x = issue, y = records_flagged)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  labs(
    x = "Flag Type",
    y = "Number of Records",
    title = "Data Quality Flags in Kenya's GBIF Records"
  )
```

## Discovery 4: Environmental bias is systematic

I compared the environmental conditions at sampled locations against Kenya’s full environmental space. Every variable — elevation, temperature, precipitation — showed **statistically significant bias**.

Sampling concentrates in **mid-elevation, moderate-climate areas near cities and roads**. The environmental extremes — hot lowlands, wet forests, alpine zones — are underrepresented.

### Interactive Comparison: Sampled vs Available Environments

```{r environmental-bias}
#| fig-cap: "Environmental bias assessment. Red = sampled locations, Grey = available environmental space across Kenya. Significant divergence indicates systematic sampling bias."
#| fig-height: 8

# Prepare data for plotly
env_long <- env_comparison %>%
  pivot_longer(cols = c(elevation, temperature, precipitation),
               names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(
    variable_label = case_when(
      variable == "elevation" ~ "Elevation (m)",
      variable == "temperature" ~ "Mean Annual Temperature (°C × 10)",
      variable == "precipitation" ~ "Annual Precipitation (mm)"
    )
  )

density_sampled <- env_long %>%
  filter(type == "sampled") %>%
  group_by(variable, variable_label) %>%
  summarise(
      density_data = list(density(value, na.rm = TRUE)),
      .groups = "drop"
  )

density_available <- env_long %>%
  filter(type == "available") %>%
  group_by(variable, variable_label) %>%
  summarise(
      density_data = list(density(value, na.rm = TRUE)),
      .groups = "drop"
  )

# Build plotly object
plots <- lapply(unique(env_long$variable), function(var) {
  var_label <- env_long %>%
    filter(variable == var) %>%
    pull(variable_label) %>%
    unique()

  d_s <- density_sampled$density_data[[which(density_sampled$variable == var)]]
  d_a <- density_available$density_data[[which(density_available$variable == var)]]

  plot_ly() %>%
    add_lines(
      x = d_a$x, y = d_a$y,
      name = "Available space",
      type = "scatter",
      mode = "lines",
      line = list(color = 'rgba(128, 128, 128, 0.7)', width = 2),
      fill = 'tozeroy',
      fillcolor = 'rgba(128, 128, 128, 0.3)',
      hovertemplate = paste0('<b>Available</b><br>',
                             var_label, ': %{x:.1f}<br>',
                             'Density: %{y:.4f}<extra></extra>')
    ) %>%
    add_lines(
      x = d_s$x, y = d_s$y,
      name = "Sampled locations",
      type = "scatter",
      mode = "lines",
      fill = 'tozeroy',
      fillcolor = 'rgba(255, 0, 0, 0.4)',
      line = list(color = 'rgb(255, 0, 0)', width = 2),
      hovertemplate = paste0('<b>Sampled Locations</b><br>',
                             var_label, ': %{x:.1f}<br>',
                             'Density: %{y:.4f}<extra></extra>')
    ) %>%
    layout(
      title = list(text = var_label, font = list(size = 14)),
      xaxis = list(title = var_label),
      yaxis = list(title = "Density"),
      hovermode = 'closest',
      showlegend = TRUE,
      legend = list(x = 0.7, y = 0.95)
    )
})

subplot(plots, nrows = length(plots), shareX = FALSE, shareY = FALSE, margin = 0.05) %>%
  layout(
    title = list(
      text = "Environmental Bias: Sampled vs Available Space",
      font = list(size = 16, color = '#333')
    ),
    showlegend = TRUE,
    height = 700
  )
```

## What This Means for Kenya

These findings aren't cause for despair. They're a **roadmap**.

**The gaps are now mapped.** We know precisely which regions and taxa need attention. Future sampling can be strategic rather than opportunistic.

**The biases are quantified.** Researchers can incorporate bias corrections into models. Conservation planners can distinguish between "species absent" and "species unsampled."

**A baseline exists.** Kenya's Environment Institute, conservation organisations, and research institutions now have documented metrics to measure progress against.

**Opportunities abound.** Citizen science programmes can target undersampled areas. University research can prioritise neglected taxa. International collaborations can address the data gaps systematically.

From a national perspective, this GBIF-based baseline is **complementary** to internal wildlife and biodiversity monitoring systems. It highlights where Kenya’s global data footprint is strong or weak, even if internal datasets (for example, in KWS, WRTI or museums) are richer than what is currently visible here.

### How this connects to Kenya's Vision 2030 and national wildlife data systems

Kenya's **Vision 2030** explicitly calls for identifying, mapping and documenting wildlife hotspots, strengthening connectivity between them to enhance the ecological integrity of habitats, refurbishing and equipping wildlife research stations, enhancing ecological monitoring programmes in all protected areas, and developing a **national wildlife research, information and database** at Kenya Wildlife Service (KWS) headquarters.

In practice, a number of complementary initiatives are moving in this direction: the Kenya Biodiversity Information System and CBD Clearing-House Mechanism, the Biodiversity Atlas of Kenya and associated web portal, internal wildlife information systems within KWS and the Wildlife Research and Training Institute (WRTI), and emerging national wildlife data portals that aim to integrate data across agencies.

This GBIF-based assessment does not attempt to duplicate those systems. Instead, it **quantifies what is currently visible in the global open data infrastructure** and highlights where that view is thin or biased. In other words, it shows part of the gap between **what Kenya already knows internally about its biodiversity** and **what the world can currently see and reuse**—a gap that Vision 2030 and ongoing national data initiatives are explicitly trying to close.

## The Bigger Picture

Similar country- and global-scale assessments using GBIF and other digital accessible information have found **very similar patterns**. Meyer et al. (2015) identify large gaps and biases in global vertebrate point records, especially in emerging economies, and show that data completeness is strongly related to proximity to researchers and participation in data-sharing networks. Beck et al. (2014) demonstrate that spatial bias in GBIF can substantially distort species distribution models, and that subsampling to reduce bias can actually improve predictive performance even with fewer records. Oliver et al. (2021) develop global indicators of biodiversity data coverage and sampling effectiveness and find that, despite rapid growth in records, strong geographic and taxonomic biases persist. Hughes (2021) synthesises global patterns of taxonomic, spatial and environmental bias across hundreds of millions of records and concludes that sampling biases systematically shape our view of the natural world.

Kenya isn't unique. These patterns — accessibility bias, taxonomic bias, environmental bias — appear in biodiversity data globally. What differs is whether we acknowledge and quantify them honestly.

I built this analysis to be **fully reproducible**. Every statistic traces to source code. The methodology transfers to any country, any taxonomic group, any observational dataset. Because transparent, reproducible science is how we move from data to knowledge.

Kenya's biodiversity is far richer than our current data reflects. Admitting what we don't know is the first step toward knowing more.

### Further reading

::: {.callout-note}
## Further reading on global coverage and bias in biodiversity data

If you want to situate Kenya's GBIF patterns in the broader literature, these are good starting points:

- **Meyer, C., Kreft, H., Guralnick, R., & Jetz, W. (2015).** Global priorities for an effective information basis of biodiversity distributions. *Nature Communications*, 6, 8221. https://doi.org/10.1038/ncomms9221  
- **Beck, J., Böller, M., Erhardt, A., & Schwanghart, W. (2014).** Spatial bias in the GBIF database and its effect on modeling species' geographic distributions. *Ecological Informatics*, 19, 10–18. https://doi.org/10.1016/j.ecoinf.2013.11.002  
- **Oliver, R. Y., Meyer, C., Ranipeta, A., Winner, K., & Jetz, W. (2021).** Global and national trends, gaps, and opportunities in documenting and monitoring species distributions. *PLOS Biology*, 19(8), e3001336. https://doi.org/10.1371/journal.pbio.3001336  
- **Hughes, A. C. (2021).** Sampling biases shape our view of the natural world. *Ecography*, 44(9), 1259–1269. https://doi.org/10.1111/ecog.05926  
:::

---

📖 **Full interactive analysis with complete methodology and code:**
[Comprehensive Data Quality Assessment and Bias Analysis of GBIF Biodiversity Data in Kenya](kenya_gbif_bias_assessment.html)

---

*This independent analysis was conducted to advance biodiversity data science methods and establish a baseline for Kenya's GBIF data quality. Feedback and collaboration welcome.*

::: {.callout-note}
## About This Analysis
This work was produced by [Kwiz Computing Technologies](https://kwizresearch.com), a data science consultancy specialising in R programming, environmental data science, and reproducible research workflows.
:::
